DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2017-01-13 2015CompleteDerivation/completeDerivation.tex
Adding comments from Lydia

It should be clear by now that the case of discrete topology is qualitatively different from the standard topology for $\mathbb{R}^n$. The notion that the continuous case is a limit of the discrete case, that it's ``like the discrete but with more points", leads in this case to erroneous intuition. The key question is: can we physically distinguish an isolated state? Can we have an outcome associated with only one element? The answer is no with the standard topology on $\mathbb{R}^n$. The consequence is that when we define the measure $\mu$ for the cardinality of continuous states, we assign a finite value to a finite region, and we assign zero measure to a single state. That is what gave us integration, densities and, ultimately, conjugate variables. If we were to use the discrete topology on $\mathbb{R}^n$, if we were to assume we can identify single states as we do for countable states, then we would assign measure one to each state and infinity to a finite region. We would not have integration, just a simple sum. Our distribution would not be a density and there would be no justification for conjugate variables. A finite distribution could only distribute a finite amount of material in a finite number of states.

@@

We can capture the above discussion by stating that the complete state space $\bar{\mathcal{S}}$ is a hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. And if we set $q^\alpha = (t, q^i)$ and $p_\alpha = (-E, p_i)$, we can write $\omega \equiv \sum dq^\alpha \wedge dp_\alpha$. This is the manifold that allows us to define time-and-state-variable-invariant densities. The symplectic form $\omega$ allows us to measure the cardinality of possibilities as in the time independent case.

@@

	We claim there exists a homeomorphism between $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ and a hyper-surface of $T\mathcal{M}$. Let $\mathcal{s} \in \bar{\mathcal{S}}$ be the particle state for a minimal classical material. Let $x^\alpha(s)$ be the evolution associated with that state. Let $t(\mathcal{s})$ be the instant of time at which $\mathcal{s}$ is defined. $q^\alpha(\mathcal{s}) \equiv x^\alpha(s(t(\mathcal{s})))$ are state variables as they are real function of the state. $q^\alpha$ is differentiable therefore $d_s x^\alpha$ exists. Suppose $d_s x^\alpha$ is not part of the state. Then $d_s x^\alpha = d_s q^\alpha$ must be determined by the law of evolution. $x^\alpha(s)$ is determined by the law of evolution therefore $q^\alpha$ are the only state variables needed from the trajectory. As the material is minimal, $q^\alpha$ determines the full state $\mathcal{s}$. This is a contradiction: $\mathcal{s}$ is not identified by just $q^\alpha$. $u^\alpha=d_s x^\alpha$ is part of the state. $p_\alpha = p_\alpha(x^\alpha, u^\alpha)$. The map $(x^\alpha, u^\alpha) \mapsto (q^\alpha, p_\alpha)$ is continuous by \ref{prop:continuity} and is bijective as the material is minimal. It is a homemorphism between $T^*\mathcal{M}$ and $T\mathcal{M}$. $\bar{\mathcal{S}}$ is a hyper-surface of $T^*\mathcal{M}$ therefore its image is a hyper-surface of $T\mathcal{M}$.

@@

This also allows us to give $c$ a more insightful physical meaning. It is not a velocity: the metric is defined at each point without motion. It is the constant of proportionality that maps a range of initial conditions from the time variable $t$ to the space variable $x^i$. For example, once the SI is adopted, $1$km would correspond to a thousand times more possible states than $1$m, and, in the same way, $1$sec would correspond to $299,792,458$ more possible states than $1$m. Therefore when we write $x^\alpha = (ct, x^i)$ or $p_\alpha = (- E/c, p_i)$ we are really expressing everything in quantities that are proportional to the cardinality of states and possibilities. These coordinates make the bookkeeping of number of states much easier which, as we saw, is the only thing that Hamiltonian and Lagrangian mechanics are doing.

This also allows us to understand conceptually why the speed of deterministic and reversible motion cannot be greater than $c$. Consider the trajectory of a point particle. Over an interval $dt$ it will travel $dx$ in space. Since the motion is deterministic and reversible, and therefore continuous, it will travel through all the states defined in the interval $dx$. It will also travel through every moment in time, which corresponds to $cdt$ possibilities. As it moves, time will always change but not necessarily space. The number of states traveled in space cannot exceed the number of states traveled in time. Therefore we have:
\begin{align*}
\frac{|n_x(dx)|}{|n_t(dt)|} &=  \frac{|dx|}{|c dt|} \leq 1 \\
\frac{|dx|}{|dt|} & \leq c \\
\end{align*}
Note, though, that this constraint applies only to deterministic and reversible motion. It was under that premise that we were able to show that the trajectories are continuous, that we could define $\omega$ and therefore $g$. Non-local quantum effects, such as the ones featured in EPR\cite{EPR} experiments, are not examples of deterministic and reversible motion. Therefore they do not have anything to violate: the maximum speed does not apply to them. On the other hand, only deterministic and reversible motion can carry information and can establish before/after relationships, therefore everything remains consistent.

>>>>

It should be clear by now that the case of discrete topology is qualitatively different from the standard topology for $\mathbb{R}^n$. The notion that the continuous case is a limit of the discrete case, that it's ``like the discrete but with more points", leads in this case to erroneous intuition. The key question is: can we physically distinguish an isolated state? Can we have an outcome associated with only one element? The answer is no with the standard topology on $\mathbb{R}^n$. The consequence is that when we define the measure $\mu$ for the count of continuous states, we assign a finite value to a finite region, and we assign zero measure to a single state. That is what gave us integration, densities and, ultimately, conjugate variables. If we were to use the discrete topology on $\mathbb{R}^n$, if we were to assume we can identify single states as we do for countable states, then we would assign measure one to each state and infinity to a finite region. We would not have integration, just a simple sum. Our distribution would not be a density and there would be no justification for conjugate variables. A finite distribution could only distribute a finite amount of material in a finite number of states.

@@

We can capture the above discussion by stating that the complete state space $\bar{\mathcal{S}}$ is a hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. And if we set $q^\alpha = (t, q^i)$ and $p_\alpha = (-E, p_i)$, we can write $\omega \equiv \sum dq^\alpha \wedge dp_\alpha$. This is the manifold that allows us to define time-and-state-variable-invariant densities. The symplectic form $\omega$ allows us to measure the count of possibilities as in the time independent case.

@@

	We claim there exists a homeomorphism between $\bar{\mathcal{S}} \subset T^*\mathcal{M}$ and a hyper-surface of $T\mathcal{M}$. Let $\mathcal{s} \in \bar{\mathcal{S}}$ be the particle state for a minimal classical material. Let $x^\alpha(s)$ be the evolution associated with that state. Let $t(\mathcal{s})$ be the instant of time at which $\mathcal{s}$ is defined. $q^\alpha(\mathcal{s}) \equiv x^\alpha(s(t(\mathcal{s})))$ are state variables as they are real function of the state. $q^\alpha$ is differentiable therefore $d_s x^\alpha$ exists. Suppose $d_s x^\alpha$ is not part of the state. Then $d_s x^\alpha = d_s q^\alpha$ must be determined by the law of evolution. $x^\alpha(s)$ is determined by the law of evolution therefore $q^\alpha$ are the only state variables needed from the trajectory. As the material is minimal, $q^\alpha$ determines the full state $\mathcal{s}$. This is a contradiction: $\mathcal{s}$ is not identified by just $q^\alpha$. Therefore $u^\alpha=d_s x^\alpha$ is part of the state. $p_\alpha = p_\alpha(x^\alpha, u^\alpha)$. The map $(x^\alpha, u^\alpha) \mapsto (q^\alpha, p_\alpha)$ is continuous by \ref{prop:continuity} and is bijective as the material is minimal. It is a homemorphism between $T^*\mathcal{M}$ and $T\mathcal{M}$. $\bar{\mathcal{S}}$ is a hyper-surface of $T^*\mathcal{M}$ therefore its image is a hyper-surface of $T\mathcal{M}$.

@@

This also allows us to give $c$ a more insightful physical meaning. It is not a velocity: the metric is defined at each point without motion. It is the constant of proportionality that maps a range of initial conditions from the time variable $t$ to the space variable $x^i$. For example, once the SI is adopted, $1$km would correspond to a thousand times more possible states than $1$m, and, in the same way, $1$sec would correspond to $299,792,458$ more possible states than $1$m. Therefore when we write $x^\alpha = (ct, x^i)$ or $p_\alpha = (- E/c, p_i)$ we are really expressing everything in quantities that are proportional to the count of states and possibilities. These coordinates make the bookkeeping of number of states much easier which, as we saw, is the only thing that Hamiltonian and Lagrangian mechanics are doing.

This also allows us to understand conceptually why the speed of deterministic and reversible motion cannot be greater than $c$. Consider the trajectory of a point particle. Over an interval $dt$ it will travel $dx$ in space. Since the motion is deterministic and reversible, and therefore continuous, it will travel through all the states defined in the interval $dx$. It will also travel through every moment in time, which corresponds to $cdt$ possibilities. As it moves, time will always change but not necessarily space. The number of states traveled in space cannot exceed the number of states traveled in time. Therefore we have:
\begin{align*}
\frac{|n_x(dx)|}{|n_t(dt)|} &=  \frac{|dx|}{|c dt|} \leq 1 \\
\frac{|dx|}{|dt|} & \leq c \\
\end{align*}
Note, though, that this constraint applies only to deterministic and reversible motion. It was under that premise that we were able to show that the trajectories are continuous, that we could define $\omega$ and therefore $g$. Non-local quantum effects, such as the ones featured in EPR\cite{EPR} experiments, are not examples of deterministic and reversible evolution. Therefore they do not have anything to violate: the maximum speed does not apply to their ``motion".\footnote{The very notion of velocity becomes ill defined in such cases precisely because we do not have a unique well-defined evolution.} On the other hand, only deterministic and reversible motion can carry information and can establish before/after relationships, therefore everything remains consistent.
